Orienting the causal relationship between imprecisely measured traits using GWAS summary data

### 2017-07-27 docs/manuscript.tex
submitted revision

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using GWAS summary data}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{24 July 2017}

@@

MR can be applied to the same data - phenotypic measures of the exposure
and the outcome variables and a genetic instrument for the exposure -
but the genetic instrument is employed in a subtly different manner.
Here the SNP is used as a surrogate for the exposure. Assuming the SNP
associates with the outcome only through the exposure, the causal effect
of the exposure on the outcome can be estimated by scaling the
association between the SNP and the outcome by the association between
the SNP and the exposure. Though difficult to test empirically, this
assumption can be relaxed in various ways when multiple instruments are
available for a putative exposure (24,25) and a number of sensitivity
tests are now available to improve reliability (26). Additionally, if
valid genetic instruments are known for both traits of interest then MR
can be performed in both directions (bi-directional MR), testing the
influence of one trait on the other and vice versa, to infer the causal
direction between the two phenotypes.

@@

Assuming biological knowledge of genetic associations can be problematic
because if there exists a putative association between two variables,
with the SNP being robustly associated with each, it can be difficult to
determine which of the two variables is subject to the primary effect of
the SNP (i.e.~for which of the two variables is the SNP a valid
instrument? Figure 1). By definition, we expect that if the association
is causal then a SNP for the exposure will be associated with the
outcome, such that if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Genome-wide
association studies (GWASs) that identify genetic associations for
complex traits are, by design, hypothesis free and agnostic of genomic
function, and it often takes years of follow up studies to understand
the biological nature of a putative GWAS hit (27). If multiple
instruments are available for an hypothesised exposure, which is
increasingly typical for complex traits that are analysed in large GWAS
consortia, then techniques can be applied to mitigate these issues (16).
But these techniques cannot always be applied in the case of determining
causal directions between 'omic measures where typically only one
cis-acting SNP is known. For example if a DNA methylation probe is
associated with expression of an adjacent gene, then is a cis-acting SNP
an instrument for the DNA methylation level, or the gene expression
level (Figure 1)?

MR has some important advantages over the mediation-based approaches.
First, the mediation-based approaches require that the exposure, outcome
and instrumental variables are all measured in the same data, whereas
recent extensions to MR circumvent this requirement, allowing causal
inference to be drawn when exposure variables and outcome variables are
measured in different samples (28). This has the crucial advantage of
improving statistical power by allowing analysis in much larger sample
sizes, and dramatically expands the breadth of possible phenotypic
relationships that can be evaluated (26). Second, the mediation-based
approach of adjusting the outcome for the exposure to nullify the
association between the SNP and the outcome is affected by unmeasured
confounding of the exposure and outcome. This is because adjusting the
outcome by the exposure induces a collider effect between the SNP and
outcome (29), and in order to fully abrogate this association one must
also adjust for all (hidden or otherwise) confounders. MR does not
suffer from this problem because it does not test for association
through adjustment. Third, when MR assumptions are satisfied the method
is robust to there being measurement error in the exposure variable
(30). Indeed instrumental variable (IV) analysis was in part initially
introduced as a correction for measurement error in the exposure (31),
whereas it has been noted that both classic mediation-based analyses
(13,14,32,33) and mediation-based methods that use instrumental
variables (34,35) are prone to be unreliable in its presence.

Using theory and simulations we show how non-differential measurement
error in phenotypes can lead to unreliable causal inference in the
mediation-based CIT method. Though we only examine the CIT method in
detail, we believe that attempting to adjust for mediating variables to
make causal inference is susceptible to problems, which can be
generalised to other mediation-based methods. We then present an
extension to MR that allows researchers to ascertain the causal
direction of an association even when the biology of the instruments are
not fully understood, and also a metric to evaluate the sensitivity of
the result of this extension to measurement error. Finally, to
demonstrate the potential impact of measurement error we apply this
method to infer the direction of causation between DNA methylation
levels and gene expression levels. Our analyses highlight that because
these different causal inference techniques have varying strengths and
weaknesses, triangulation of evidence from as many sources as possible
should be practiced in causal inference (36).

@@

Measurement error of an exposure can be modeled as a transformation of
the true value (\(x\)) that leads to the observed value, \(x_o = f(x)\).
For example, following Pierce and VanderWeele (30) we can define

@@

In this study we assume there is no measurement error in the SNP. Common
genetic variants are typically less susceptible to measurement error due
to strict quality control procedures prior to genome wide association
studies. Any non-differential measurement error that might be present
(either because the SNP is poorly typed or because the SNP is not in
complete linkage disequilibrium with the causal variant) will reduce
power in MR but will not incur bias (3,13,30). We also assume that
measurement error in the exposure and the outcome are uncorrelated.

@@

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value, though we stress that for real analyses it is
not good practice to rely on p-values for making causal inference, nor
is it reliable to depend on arbitrary significance thresholds (37).

@@

Two stage least squares (2SLS) is a commonly used technique for
performing MR when the exposure, outcome and instrument data are all
available in the same sample. A p-value for this test, \(p_{MR}\), was
obtained using the \(systemfit\) function in the R package
\(R/systemfit\) (38). Note that the value of \(p_{MR}\) is identical
when using the same genetic variant to instrument the influence of the
exposure \(x\) on the outcome \(y\), or erroneously, instrumenting the
outcome \(y\) on the exposure \(x\).

@@

To infer the direction of causality it is desirable to know which of the
variables, \(x\) or \(y\), is being directly influenced by the
instrument \(g\). This can be achieved by assessing which of the two
variables has the biggest absolute correlation with \(g\) (Appendix 2),
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population (39). It is calculated as

@@

Note that the same correlation test approach can be applied to a
two-sample MR setting. Two-sample MR refers to the case where the
SNP-exposure association and SNP-outcome association are calculated in
different samples (e.g.~from publicly available summary statistics
(26,28)). Here the Steiger test of two independent correlations can be
applied where.

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package \emph{R/psych} (40).

@@

Two sample MR (28) was performed using the summary statistics for
genetic influences on gene expression and DNA methylation. To do this we
obtained a list of 458 gene expression - DNA methylation associations as
reported in Shakhbazov et al (41). These were filtered to be located on
the same chromosome, have robust correlations after correcting for
multiple testing, and to share a SNP that had a robust cis-acting effect
on both the DNA methylation probe and the gene expression probe. Because
only summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

All analysis was performed using the R programming language (42) and
code is made available at
\url{https://github.com/explodecomputer/causal-directions} and
implemented in the MR-Base (\url{http://wwww.mrbase.org}) platform (26).

@@

If we do not know whether the SNP \(g\) has a primary influence on \(x\)
or \(y\) then CIT can attempt to infer the causal direction. Though
bi-directional MR can be used to orient causal directions, this requires
knowledge of a valid instrument for each trait, and we were motivated to
develop the MR Steiger method that could operate on summary data to
orient the direction of causality using the same conditions as the CIT,
where the underlying biology of a single SNP is not fully understood. We
go on to explore the scenarios in which the method is likely to return
the correct or incorrect causal directions.

We performed simulations to compare the power and type 1 error rates of
MR and CIT in detecting a causal association between simulated variables
under different levels of imprecision simulated in the exposure.
Comparing the performance of methods with different sets of assumptions
can be difficult, but a basic comparison is shown in Figure 3. We
observe that the CIT is more conservative under the null model of no
association owing to the omnibus test statistic comprising several
statistical tests. The FDR using a p-value threshold of 0.05 appears to
be close to 0, whereas for the MR Steiger method the FDR is around 0.05.
Using the same p-value thresholds to declare significance in the
non-null simulations, the general trend appears to be that the CIT power
reduces as measurement error in the exposure increases faster than that
of the MR Steiger method.

@@

Unmeasured confounding between the exposure and outcome can also give
rise to problems with the MR Steiger approach (Appendix 3). The
relationship between unmeasured confounding and causal orientation is
complex across the parameter space of possible confounding values
(Supplementary Figure 1: Graph representing the unmeasured confounding
parameters that will lead to the MR Steiger test returning the wrong
causal direction. Columns of boxes represent different signed values of
the observational variance explained between \(x\) and \(y\)
(\(R^2_{xy}\)).). Based on the range of parameter values that we
explored, when the magnitude of the observational variance explained
between the exposure and the outcome is below 0.2 the MR Steiger method
is unlikely to return the incorrect causal direction due to unmeasured
confounding.

@@

We performed two sample MR (28) for each association in the direction of
causality inferred by the Stieger test. We observed that the sign of the
MR estimate was generally in the same direction as the Pearson
correlation coefficient reported by Shakhbazov et al (41) (Figure 6b).
There was a moderate correlation between the absolute magnitudes of the
causal correlation and the observational Pearson correlation (r = 0.45).
Together these inferences suggest that even in estimating associations
between `omic' variables, which are considered to be low level
phenotypes, it is important to use causal inference methods over
observational associations to infer causal effect sizes.

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.61 (95\%
CI 0.36 - 1.03), Figure 6c), suggesting that reducing methylation levels
at a controlling CpG typically leads to increased gene expression
levels, consistent with expectation (43).

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis are relatively well explored (44). Our analysis extends this to
related methods such as CIT that are used in predominantly 'omic data.
These methods are indeed susceptible to the same problem as standard
mediation based analysis, and specifically we show that as measurement
error in the (true) exposure variable increases, CIT is likely to have
reduced statistical power, and liable to infer the wrong direction of
causality. We also demonstrate that, though unintuitive, increasing
sample size does not resolve the issue, rather it leads to more extreme
p-values for the model that predicts the wrong direction of causality.

@@

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The inferred causal direction is
heavily influenced by how much measurement error is present in the
different assaying platforms. For example, if DNA methylation measures
typically have higher measuremet error than gene expression measures
then our analysis suggests that DNA methylation levels would be more
often the causal factor in the association. Indeed, previous studies
which have evaluated measurement error in these platforms do support
this position (45,46), though making strong conclusions for this
analysis is difficult because measurement error is likely to be study
specific. We also haven't accounted for the influence of winner's curse,
which can inflate estimates of the variance explained by SNPs, with
higher inflation expected amongst lower powered studies. Using p-values
for genetic associations from replication studies will mitigate this
problem.

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. However,
if there is only a single instrument it is difficult to separate between
the two competing models of \(g\) instrumenting a trait which causes
another trait, and \(g\) having pleiotropic effects on both traits
independently (47). Under certain conditions of measurement error the
CIT test can distinguish these models. We also note that it is
straightforward to extend the MR Steiger approach to multiple
instruments, requiring only that the total variance explained by all
instruments be calculated under the assumption that they are
independent. Multiple instruments can indeed help to distinguish between
the causal and pleiotropic models, for example by evaluating the
proportionality of the SNP-exposure and SNP-outcome effects (16).
Additionally, if there is at least one instrument for each trait then
bi-directional MR can offer solutions to inferring the causal direction
(16,48,49). We restricted the simulations to evaluating the causal
inference between quantitative traits, but it is possible that the
analysis could be extended to binary traits by using the genetic
variance explained on the liability scale, taking into account the
population prevalence (50). However, our analysis goes beyond many
previous explorations of measurement error by assessing the impacts of
both imprecision (noise) and linear transformations of the true variable
on causal inference.

Our new method attempts to infer causal directions under the assumption
that horizontal pleiotropy (the influence of the instrument on the
outcome through a mechanism other than the exposure) is not present.
Recent method developments in MR (24,25) have focused on accounting for
the issues that horizontal pleiotropy can introduce when multiple
instruments are available, but how they perform in the presence of
measurement error remains to be explored. An important advantage that MR
confers over most mediation based analysis is that it can be performed
in two samples, which can considerably improve power and expand the
scope of analysis. However, whether there is a substantive difference in
two sample MR versus one sample MR in how measurement error has an
effect is not yet fully understood. We have also assumed no measurement
error in the genetic instrument, which is not unreasonable given the
strict QC protocols that ensure high quality genotype data is available
to most studies. We have restricted the scope to only exploring
non-differential measurement error and avoided the complications
incurred if measurement error in the exposure and outcome is correlated.
We have also not addressed other issues pertaining to instrumental
variables which are relevant to the question of instrument-exposure
specification. One such problem is exposure misspecification, for
example an instrument could associate with several closely related
putative outcomes, with only one of them actually having a causal effect
on the outcome. This problem has shown to be the case for SNPs
influencing different lipid fractions, for example (51,52).

Mediation based network approaches, that go beyond analyses of two
variables, are very well established (35) and have a number of
extensions that make them valuable tools, including for example network
construction. But because they are predicated on the basic underlying
principles of mediation they are liable to suffer from the same issues
of measurement error. Recent advances in MR methodology, for example
applying MR to genetical genomics (53), multivariate MR (52) and
mediation through MR (54--56) may offer more robust alternatives for
these more complicated problems.

@@

Supplementary Figure 1 shows the relationship between the magnitude of
the correlations between \(x\), \(y\) and the confounder \(u\) for a
range of \(\beta_{xy} = (-2,2)\), \(\beta_{gx} = 0.1\) and a range of
confounder effects. The pattern of results were similar for different
values of \(\beta_{gx}\). We note that in most cases for the parameter
values explored, where the observational \(\hat{R}^2_{xy}\) is less than
0.2, unmeasured confounding will not incur the wrong causal direction in
the MR Steiger test.

@@

\hypertarget{ref-Claussnitzer2015}{}
27. Claussnitzer M, Dankel SN, Kim K-H, Quon G, Meuleman W, Haugen C, et
al. FTO Obesity Variant Circuitry and Adipocyte Browning in Humans. The
New England journal of medicine. 2015;373(10):895--907.

\hypertarget{ref-Pierce2013}{}
28. Pierce BL, Burgess S. Efficient design for Mendelian randomization
studies: subsample and 2-sample instrumental variable estimators.
American journal of epidemiology {[}Internet{]}. 2013
Oct;178(7):1177--84. Available from:
\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3783091\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3783091\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}

\hypertarget{ref-Hernan2004}{}
29. Hernán MA, Hernández-Díaz S, Robins JM. A structural approach to
selection bias. Epidemiology (Cambridge, Mass) {[}Internet{]}. 2004
Sep;15(5):615--25. Available from:
\url{http://www.ncbi.nlm.nih.gov/pubmed/15308962}

\hypertarget{ref-Pierce2012}{}
30. Pierce BL, VanderWeele TJ. The effect of non-differential
measurement error on bias, precision and power in Mendelian
randomization studies. International Journal of Epidemiology
{[}Internet{]}. 2012 Oct;41(5):1383--93. Available from:
\url{https://academic.oup.com/ije/article-lookup/doi/10.1093/ije/dys141}

\hypertarget{ref-Ashenfelter1994}{}
31. Ashenfelter O, Krueger AB. Estimates of the Economic Return to
Schooling from a New Sample of Twins. The American Economic Review.
1994;84(5):1157--73.

\hypertarget{ref-Nagarajan2013}{}
32. Nagarajan R, Scutari M. Impact of noise on molecular network
inference. PloS one {[}Internet{]}. 2013 Jan;8(12):e80735. Available
from:
\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3855153\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3855153\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}

\hypertarget{ref-Shpitser2010}{}
33. Shpitser I, VanderWeele T, Robins J. On the validity of covariate
adjustment for estimating causal effects. Proceedings of the Twenty
Sixth Conference on Uncertainty in Artificial Intelligence (UAI-10).
2010;527--36.

\hypertarget{ref-Wang2015}{}
34. Wang L, Michoel T. Detection of regulator genes and eQTLs in gene
networks. arXiv {[}Internet{]}. 2015 Dec;arXiv:1512. Available from:
\url{http://arxiv.org/abs/1512.05574}

\hypertarget{ref-Lagani2015}{}
35. Lagani V, Triantafillou S, Ball G, Tegner J, Tsamardinos I.
Probabilistic Computational Causal Discovery for Systems Biology. In:
Uncertainty in biology: A computational modeling approach
{[}Internet{]}. Springer; 2015. p. 47. Available from:
\href{https://books.google.com/books?id=8SLUCgAAQBAJ\%7B/\&\%7Dpgis=1}{https://books.google.com/books?id=8SLUCgAAQBAJ\{\textbackslash{}\&\}pgis=1}

>>>>

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using GWAS summary data}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{27 July 2017}

@@

MR can be applied to the same data - phenotypic measures of the exposure
and the outcome variables and a genetic instrument for the exposure -
but the genetic instrument is employed in a subtly different manner.
Here the SNP is used as a surrogate for the exposure. Assuming the SNP
associates with the outcome only through the exposure, the causal effect
of the exposure on the outcome can be estimated by scaling the
association between the SNP and the outcome by the association between
the SNP and the exposure. Though difficult to test empirically, this
assumption can be relaxed in various ways when multiple instruments are
available for a putative exposure (24,25) and a number of sensitivity
tests are now available to improve reliability (26). Additionally, if
valid genetic instruments are known for both traits of interest then MR
can be performed in both directions (bi-directional MR), testing the
influence of one trait on the other and vice versa, to infer the causal
direction between the two phenotypes (27,28).

@@

Assuming biological knowledge of genetic associations can be problematic
because if there exists a putative association between two variables,
with the SNP being robustly associated with each, it can be difficult to
determine which of the two variables is subject to the primary effect of
the SNP (i.e.~for which of the two variables is the SNP a valid
instrument? Figure 1). By definition, we expect that if the association
is causal then a SNP for the exposure will be associated with the
outcome, such that if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Genome-wide
association studies (GWASs) that identify genetic associations for
complex traits are, by design, hypothesis free and agnostic of genomic
function, and it often takes years of follow up studies to understand
the biological nature of a putative GWAS hit (29). If multiple
instruments are available for an hypothesised exposure, which is
increasingly typical for complex traits that are analysed in large GWAS
consortia, then techniques can be applied to mitigate these issues (16).
But these techniques cannot always be applied in the case of determining
causal directions between 'omic measures where typically only one
cis-acting SNP is known. For example if a DNA methylation probe is
associated with expression of an adjacent gene, then is a cis-acting SNP
an instrument for the DNA methylation level, or the gene expression
level (Figure 1)?

MR has some important advantages over the mediation-based approaches.
First, the mediation-based approaches require that the exposure, outcome
and instrumental variables are all measured in the same data, whereas
recent extensions to MR circumvent this requirement, allowing causal
inference to be drawn when exposure variables and outcome variables are
measured in different samples (30). This has the crucial advantage of
improving statistical power by allowing analysis in much larger sample
sizes, and dramatically expands the breadth of possible phenotypic
relationships that can be evaluated (26). Second, the mediation-based
approach of adjusting the outcome for the exposure to nullify the
association between the SNP and the outcome is affected by unmeasured
confounding of the exposure and outcome. This is because adjusting the
outcome by the exposure induces a collider effect between the SNP and
outcome (31), and in order to fully abrogate this association one must
also adjust for all (hidden or otherwise) confounders. MR does not
suffer from this problem because it does not test for association
through adjustment. Third, when MR assumptions are satisfied the method
is robust to there being measurement error in the exposure variable
(32). Indeed instrumental variable (IV) analysis was in part initially
introduced as a correction for measurement error in the exposure (33),
whereas it has been noted that both classic mediation-based analyses
(13,14,34,35) and mediation-based methods that use instrumental
variables (36,37) are prone to be unreliable in its presence.

Using theory and simulations we show how non-differential measurement
error in phenotypes can lead to unreliable causal inference in the
mediation-based CIT method. Though we only examine the CIT method in
detail, we believe that attempting to adjust for mediating variables to
make causal inference is susceptible to problems, which can be
generalised to other mediation-based methods. We then present an
extension to MR that allows researchers to ascertain the causal
direction of an association even when the biology of the instruments are
not fully understood, and also a metric to evaluate the sensitivity of
the result of this extension to measurement error. Finally, to
demonstrate the potential impact of measurement error we apply this
method to infer the direction of causation between DNA methylation
levels and gene expression levels. Our analyses highlight that because
these different causal inference techniques have varying strengths and
weaknesses, triangulation of evidence from as many sources as possible
should be practiced in causal inference (38).

@@

Measurement error of an exposure can be modeled as a transformation of
the true value (\(x\)) that leads to the observed value, \(x_o = f(x)\).
For example, following Pierce and VanderWeele (32) we can define

@@

In this study we assume there is no measurement error in the SNP. Common
genetic variants are typically less susceptible to measurement error due
to strict quality control procedures prior to genome wide association
studies. Any non-differential measurement error that might be present
(either because the SNP is poorly typed or because the SNP is not in
complete linkage disequilibrium with the causal variant) will reduce
power in MR but will not incur bias (3,13,32). We also assume that
measurement error in the exposure and the outcome are uncorrelated.

@@

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value, though we stress that for real analyses it is
not good practice to rely on p-values for making causal inference, nor
is it reliable to depend on arbitrary significance thresholds (39).

@@

Two stage least squares (2SLS) is a commonly used technique for
performing MR when the exposure, outcome and instrument data are all
available in the same sample. A p-value for this test, \(p_{MR}\), was
obtained using the \(systemfit\) function in the R package
\(R/systemfit\) (40). Note that the value of \(p_{MR}\) is identical
when using the same genetic variant to instrument the influence of the
exposure \(x\) on the outcome \(y\), or erroneously, instrumenting the
outcome \(y\) on the exposure \(x\).

@@

To infer the direction of causality it is desirable to know which of the
variables, \(x\) or \(y\), is being directly influenced by the
instrument \(g\). This can be achieved by assessing which of the two
variables has the biggest absolute correlation with \(g\) (Appendix 2),
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population (41). It is calculated as

@@

Note that the same correlation test approach can be applied to a
two-sample MR setting. Two-sample MR refers to the case where the
SNP-exposure association and SNP-outcome association are calculated in
different samples (e.g.~from publicly available summary statistics
(26,30)). Here the Steiger test of two independent correlations can be
applied where.

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package \emph{R/psych} (42).

@@

Two sample MR (30) was performed using the summary statistics for
genetic influences on gene expression and DNA methylation. To do this we
obtained a list of 458 gene expression - DNA methylation associations as
reported in Shakhbazov et al (43). These were filtered to be located on
the same chromosome, have robust correlations after correcting for
multiple testing, and to share a SNP that had a robust cis-acting effect
on both the DNA methylation probe and the gene expression probe. Because
only summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

All analysis was performed using the R programming language (44) and
code is made available at
\url{https://github.com/explodecomputer/causal-directions} and
implemented in the MR-Base (\url{http://wwww.mrbase.org}) platform (26).

@@

If we do not know whether the SNP \(g\) has a primary influence on \(x\)
or \(y\) then CIT can attempt to infer the causal direction. Though
bi-directional MR can be used to orient causal directions (27), this
requires knowledge of a valid instrument for each trait, and we were
motivated to develop the MR Steiger method that could operate on summary
data to orient the direction of causality using the same conditions as
the CIT, where the underlying biology of a single SNP is not fully
understood. We go on to explore the scenarios in which the method is
likely to return the correct or incorrect causal directions.

We performed simulations to compare the power and type 1 error rates of
MR and CIT in detecting a causal association between simulated variables
under different levels of imprecision simulated in the exposure.
Comparing the performance of methods with different sets of assumptions
can be difficult, but a basic comparison is shown in Figure 3. We
observe that the CIT is more conservative under the null model of no
association owing to the omnibus test statistic comprising several
statistical tests. The FDR using a p-value threshold of 0.05 appears to
be close to 0, whereas for the MR Steiger method the FDR is around 0.05.
Using the same p-value thresholds to declare significance in the
non-null simulations, the general trend appears to be that the CIT power
reduces as measurement error in the exposure increases more steeply than
that of the MR Steiger method.

@@

Unmeasured confounding between the exposure and outcome can also give
rise to problems with the MR Steiger approach (Appendix 3). The
relationship between unmeasured confounding and causal orientation is
complex across the parameter space of possible confounding values
(Supplementary Figure 1). Based on the range of parameter values that we
explored, when the magnitude of the observational variance explained
between the exposure and the outcome is below 0.2 the MR Steiger method
is unlikely to return the incorrect causal direction due to unmeasured
confounding.

@@

We performed two sample MR (30) for each association in the direction of
causality inferred by the Stieger test. We observed that the sign of the
MR estimate was generally in the same direction as the Pearson
correlation coefficient reported by Shakhbazov et al (43) (Figure 6b).
There was a moderate correlation between the absolute magnitudes of the
causal correlation and the observational Pearson correlation (r = 0.45).
Together these inferences suggest that even in estimating associations
between `omic' variables, which are considered to be low level
phenotypes, it is important to use causal inference methods over
observational associations to infer causal effect sizes.

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.61 (95\%
CI 0.36 - 1.03), Figure 6c), suggesting that reducing methylation levels
at a controlling CpG typically leads to increased gene expression
levels, consistent with expectation (45).

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis are relatively well explored (46). Our analysis extends this to
related methods such as CIT that are used in predominantly 'omic data.
These methods are indeed susceptible to the same problem as standard
mediation based analysis, and specifically we show that as measurement
error in the (true) exposure variable increases, CIT is likely to have
reduced statistical power, and liable to infer the wrong direction of
causality. We also demonstrate that, though unintuitive, increasing
sample size does not resolve the issue, rather it leads to more extreme
p-values for the model that predicts the wrong direction of causality.

@@

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The inferred causal direction is
heavily influenced by how much measurement error is present in the
different assaying platforms. For example, if DNA methylation measures
typically have higher measuremet error than gene expression measures
then our analysis suggests that DNA methylation levels would be more
often the causal factor in the association. Indeed, previous studies
which have evaluated measurement error in these platforms do support
this position (47,48), though making strong conclusions for this
analysis is difficult because measurement error is likely to be study
specific. We also haven't accounted for the influence of winner's curse,
which can inflate estimates of the variance explained by SNPs, with
higher inflation expected amongst lower powered studies. Using p-values
for genetic associations from replication studies will mitigate this
problem.

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. However,
if there is only a single instrument it is difficult to separate between
the two competing models of \(g\) instrumenting a trait which causes
another trait, and \(g\) having pleiotropic effects on both traits
independently (49). Under certain conditions of measurement error the
CIT test can distinguish these models. We also note that it is
straightforward to extend the MR Steiger approach to multiple
instruments, requiring only that the total variance explained by all
instruments be calculated under the assumption that they are
independent. Multiple instruments can indeed help to distinguish between
the causal and pleiotropic models, for example by evaluating the
proportionality of the SNP-exposure and SNP-outcome effects (16).
Additionally, if there is at least one instrument for each trait then
bi-directional MR can offer solutions to inferring the causal direction
(16,28,50). We restricted the simulations to evaluating the causal
inference between quantitative traits, but it is possible that the
analysis could be extended to binary traits by using the genetic
variance explained on the liability scale, taking into account the
population prevalence (51). However, our analysis goes beyond many
previous explorations of measurement error by assessing the impacts of
both imprecision (noise) and linear transformations of the true variable
on causal inference.

Our new method attempts to infer causal directions under the assumption
that horizontal pleiotropy (the influence of the instrument on the
outcome through a mechanism other than the exposure) is not present.
Recent method developments in MR (24,25) have focused on accounting for
the issues that horizontal pleiotropy can introduce when multiple
instruments are available, but how they perform in the presence of
measurement error remains to be explored. An important advantage that MR
confers over most mediation based analysis is that it can be performed
in two samples, which can considerably improve power and expand the
scope of analysis. However, whether there is a substantive difference in
two sample MR versus one sample MR in how measurement error has an
effect is not yet fully understood. We have also assumed no measurement
error in the genetic instrument, which is not unreasonable given the
strict QC protocols that ensure high quality genotype data is available
to most studies. We have restricted the scope to only exploring
non-differential measurement error and avoided the complications
incurred if measurement error in the exposure and outcome is correlated.
We have also not addressed other issues pertaining to instrumental
variables which are relevant to the question of instrument-exposure
specification. One such problem is exposure misspecification, for
example an instrument could associate with several closely related
putative outcomes, with only one of them actually having a causal effect
on the outcome. This problem has shown to be the case for SNPs
influencing different lipid fractions, for example (52,53).

Mediation based network approaches, that go beyond analyses of two
variables, are very well established (37) and have a number of
extensions that make them valuable tools, including for example network
construction. But because they are predicated on the basic underlying
principles of mediation they are liable to suffer from the same issues
of measurement error. Recent advances in MR methodology, for example
applying MR to genetical genomics (54), multivariate MR (53) and
mediation through MR (55--57) may offer more robust alternatives for
these more complicated problems.

@@

Supplementary Figure 1 shows the relationship between the magnitude of
the correlations between \(x\), \(y\) and the confounder \(u\) for a
range of \(\beta_{xy} = (-2,2)\), \(\beta_{gx} = 0.1\) and a range of
confounder effects. The pattern of results were similar for different
values of \(\beta_{gx}\). We note that in most cases for the parameter
values explored, where the observational absolute \(\hat{R}^2_{xy}\) is
less than 0.2, unmeasured confounding will not incur the wrong causal
direction in the MR Steiger test.

@@

\hypertarget{ref-Timpson2011}{}
27. Timpson NJ, Nordestgaard BG, Harbord RM, Zacho J, Frayling TM,
Tybjærg-Hansen a, et al. C-reactive protein levels and body mass index:
elucidating direction of causation through reciprocal Mendelian
randomization. International journal of obesity (2005) {[}Internet{]}.
2011 Feb;35(2):300--8. Available from:
\url{http://www.ncbi.nlm.nih.gov/pubmed/20714329}

\hypertarget{ref-Richmond2014}{}
28. Richmond RC, Davey Smith G, Ness AR, Hoed M den, McMahon G, Timpson
NJ. Assessing Causality in the Association between Child Adiposity and
Physical Activity Levels: A Mendelian Randomization Analysis. Ludwig DS,
editor. PLoS Medicine {[}Internet{]}. 2014 Mar;11(3):e1001618. Available
from: \url{http://dx.plos.org/10.1371/journal.pmed.1001618}

\hypertarget{ref-Claussnitzer2015}{}
29. Claussnitzer M, Dankel SN, Kim K-H, Quon G, Meuleman W, Haugen C, et
al. FTO Obesity Variant Circuitry and Adipocyte Browning in Humans. The
New England journal of medicine. 2015;373(10):895--907.

\hypertarget{ref-Pierce2013}{}
30. Pierce BL, Burgess S. Efficient design for Mendelian randomization
studies: subsample and 2-sample instrumental variable estimators.
American journal of epidemiology {[}Internet{]}. 2013
Oct;178(7):1177--84. Available from:
\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3783091\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3783091\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}

\hypertarget{ref-Hernan2004}{}
31. Hernán MA, Hernández-Díaz S, Robins JM. A structural approach to
selection bias. Epidemiology (Cambridge, Mass) {[}Internet{]}. 2004
Sep;15(5):615--25. Available from:
\url{http://www.ncbi.nlm.nih.gov/pubmed/15308962}

\hypertarget{ref-Pierce2012}{}
32. Pierce BL, VanderWeele TJ. The effect of non-differential
measurement error on bias, precision and power in Mendelian
randomization studies. International Journal of Epidemiology
{[}Internet{]}. 2012 Oct;41(5):1383--93. Available from:
\url{https://academic.oup.com/ije/article-lookup/doi/10.1093/ije/dys141}

\hypertarget{ref-Ashenfelter1994}{}
33. Ashenfelter O, Krueger AB. Estimates of the Economic Return to
Schooling from a New Sample of Twins. The American Economic Review.
1994;84(5):1157--73.

\hypertarget{ref-Nagarajan2013}{}
34. Nagarajan R, Scutari M. Impact of noise on molecular network
inference. PloS one {[}Internet{]}. 2013 Jan;8(12):e80735. Available
from:
\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3855153\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3855153\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}

\hypertarget{ref-Shpitser2010}{}
35. Shpitser I, VanderWeele T, Robins J. On the validity of covariate
adjustment for estimating causal effects. Proceedings of the Twenty
Sixth Conference on Uncertainty in Artificial Intelligence (UAI-10).
2010;527--36.
